The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2015-01-13 draft.tex
added some generate folders, some text edits for line profiles and a git ignore

\begin{abstract}
Many high-state non-magnetic cataclysmic variables (CVs) exhibit
blue-shifted absorption or P-Cygni profiles in ultraviolet
(UV) resonance lines, such as C~{\sc iv}~1550\AA. These features
imply the existence of powerful accretion disk winds in
CVs.
% and considerable effort has gone into modelling UV line formation
% in such outflows. By contrast, much less attention has been paid to
% the potential impact of these disk winds on the {\em optical} spectra
% of CVs. 
Here, we use an improved version of our Monte Carlo ionization
and radiative transfer code to investigate whether disk wind models that
produce realistic UV line profiles are also likely to generate
observationally significant recombination line and continuum emission
in the {\em optical} waveband.   
In particular, we test whether outflows may 
be responsible for the single-peaked emission line profiles often seen
in high-state CVs and for the weakness of the Balmer absorption edge
(relative to simple models of optically thick accretion disks).
We find that a standard disk wind model that is successful in
reproducing the UV spectra of CVs also leaves a noticeable imprint on
the optical spectrum, particularly for systems viewed at high
inclination. The strongest optical wind-formed recombinations lines are \ha\ and
\heiiopt\, although the outflow also produces emission in the Ly-$\alpha$
and \heiiuv\ UV lines. However, all of the optical lines produced by
this default wind model are double-peaked, and the effect of the wind
on the Balmer jump is modest. 
We demonstrate that a higher-density outflow model 
displays reasonable verisimilitude with the optical spectrum of RW Trianguli,
producing all the observed lines. In addition, this model
can produce both single-peaked \ha\ emission and a
recombination continuum sufficient to fill in the Balmer jump,
although it also predicts stronger-than-observed 
collisionally excited UV emission lines. 
Our revised model also significantly
alters the spectral shape of the continuum and displays
a strong inclination dependence in its emission properties. 
We explicitly demonstrate that the radial velocity shear 
is responsible for single-peaked emission in our models, as predicted.
On the basis of these results, we suggest that winds can 
modify, and perhaps even dominate, the line and continuum emission
from accreting systems.
% We carry out a limited exploration of
% the disk wind parameter space, focusing on those parameters that affect the
% density near the base of the wind. This shows that higher-density
% outflows can produce both single-peaked emission lines and a
% recombination continuum sufficient to fill in the Balmer
% jump, although such models also tend to predict stronger-than-observed 
% collisionally excited UV emission lines. Our main conclusion is that
% disk wind emission is likely to have a significant impact on the
% optical spectra of high-state CVs. 
% {\bf JM: This abstract is a few lines longer than I would like- 
% I think something punchier is required.}
\end{abstract}

@@

It has been known for a long time that winds emanating from the
accretion disk are important in shaping the ultraviolet (UV) spectra
of high-state CVs (Heap 1978). The most spectacular evidence for such
outflows are the P-Cygni-like profiles seen in UV resonance lines such as
\civfull\ (see e.g. Cordova \& Mason
1982\nocite{cordova1982}). Considerable effort has been spent over the
years on understanding and modelling these UV features (e.g. Drew \&
Verbunt 1985\nocite{drewverbunt1985}; Mauche \& Raymond
1987\nocite{maucheraymond1987}; Drew 1987; Shlosman \& Vitello 1993; [hereafter
SV93]\nocite{SV93}; Knigge, Woods \& Drew 1995 [hereafter
KWD95]\nocite{KWD95}; Knigge \& Drew 1997\nocite{kd1997}; 
Knigge et al. 1997\nocite{knigge1997}; Long \&
Knigge 2002 [hereafter LK02]\nocite{LK02}, Noebauer et al. 2010\nocite{noebauer};
Puebla et al. 2011\nocite{puebla2011}). The basic picture emerging from these efforts is
of a slowly accelerating, moderately collimated bipolar
outflow that carries away $\simeq 1\% - 10\%$ of the accreting
material. State-of-art-simulations of line formation in this type
of disk wind can produce UV line profiles that are remarkably similar
to observations.

@@

Murray \& Chiang (1996, 1997; hereafter MC96, MC97)\nocite{MC96, MC97} 
have shown that the presence of disk winds may
offer a natural explanation for the single-peaked optical emission lines in
high-state CVs, since they can strongly affect the radiative transfer
of line photons. Strong support for a significant wind contribution to the
optical emission lines comes from observations of eclipsing
systems. There, the single-peaked lines are often only weakly
eclipsed, and a significant fraction of the line flux remains visible
even near mid-eclipse (REF). This points to line formation in a spatially
extended region, such as a disk wind (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence for a wind contribution to the optical lines comes
from isolated observations of P-Cygni-like line profiles even in optical
lines, such as \ha\ and He \textsc{i} $\lambda5876$ \citep{RN98, kafka2004}.
{\bf JM: need to mention this later as pointed out by CK- both in discussion
and conclusion sections.}

@@

\begin{figure}	%fullpage
\centering
\includegraphics[width=0.8\textwidth]{figures/fig1.eps}
\caption{Optical spectra of three nova-like variables: 
RW Sex (top; Beuermann et al. 1992),
IX Vel (top middle; Beuermann \& Thomas 1990) 
and RW Tri in and out of eclipse (bottom two panels; Groot et al. 2004).
The data have been digitized from the respective publications and a few of the more
prominent emission and absorption lines are marked. 
These systems have inclinations of $30^\circ$, $60^\circ$ and $70^\circ$ 
respectively.
The trend of increasing Balmer line emission with inclination can be seen.
In RW Tri strong single-peaked emission in the Balmer lines is seen even
in eclipse, indicating that the lines may be formed in a spatially
extensive disk wind, and there is even a suggestion 
of a (potentially wind-formed) recombination continuum in the eclipsed
spectrum.}
\label{novalikes}
\end{figure}

@@

\py is a Monte Carlo (MC) ionization and radiative transfer code which
uses the Sobolev approximation to treat line transfer. It has already
been described extensively in LK02, SDL05 and H13, so here we provide
only a brief summary of its operation, focusing particularly on new
aspects of our implementation of macro-atoms into the code.

@@

\py operates in two distinct stages. First, the ionization state,
level populations and temperature structure are calculated. This is
done iteratively, by 
propagating several populations of MC energy quanta (`photons')
through a model wind. The geometric and kinemetic properties of the
outflow are specified on a pre-defined spatial grid. In each of these
iterations (``ionization cycles''), the code records estimators that 
characterize the radiation field in each grid cell. At the end 
of each ionization cycles, a new electron temperature is calculated
that more closely balances heating and cooling in the 
plasma. The radiative estimators and updated electron
temperature are then used to revise the ionization state of the wind,
and a new ionization cycle is started. The process is repeated until
heating and cooling are balanced throughout the wind.

@@

Lucy (2002, 2003\nocite{lucy2002, lucy2003}; hereafter L02, L03) 
has shown that it is possible to calculate the emissivity of a gas in
statistical equilibrium accurately by quantising matter into
`macro-atoms', and radiant and kinetic energy into indivisible energy
packets (r- and k- packets, respectively). His macro-atom scheme
allows for all possible transition paths from a given level and
provides a full non-local thermodynamic equilibrium (NLTE) solution
for the level populations based on MC estimators. The macro-atom
technique has already been used to model Wolf-Rayet star
winds \citep{sim2004}, AGN disk winds \citep{simlong2008, tatum2012},
supernovae \citep{kasen2006, kerzendorfsim} and YSOs (SDL05). A full 
description of the can be found in L02 and L03.

Briefly, macro-atom NLTE level populations and ionization fractions
are calculated by solving the statistical equilbrium equations between
each pair of levels. For example, the bound-bound de-excitation rate,
${\cal R}_{ul}$, in an ion is given by  
\begin{equation}
{\cal R}_{ul} = \beta_{lu} A_{ul} n_u + \beta_{lu} B_{ul} n_u J_{lu} +
C_{ul} n_u n_e,\label{eq:nlte_rul},
\end{equation}
where $u$ and $l$ denote the upper and lower levels, $C$ represents the
collisional rate coefficients and $A$ and $B$ represent the Einstein
coefficients. The quantity $\beta_{lu}$ is the probability that a
given line photon will escape the Sobolev region. Finally, $J$ is the
MC estimator for the mean intensity in the line, as recorded during
the photon propagation. The corresponding excitation rate is then 
\begin{equation}
{\cal R}_{lu} = \beta_{lu} B_{lu} n_l J_{lu} + C_{lu} n_l n_e,
\end{equation}
where we implicitly include a correction for stimulated emission in
our definition of the estimate for $J_{ul}$.

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/acc_law.eps}
\caption{
The adopted poloidal velocity law for various values of the
acceleration exponent, $\alpha$.} 
\label{acc_law}
\end{figure}

@@

\py has has some flexibility when treating the accretion 
disk as a source of photons. The disk is broken down into annuli 
such that each annulus contributes an equal amount to the bolometric
luminosity. We take the disk to be geometrically thin, but optically
thick and thus adopt the temperature profile of a standard
\cite{shakurasunyaev1973} $\alpha$-disk. An annulus can then
be treated either as a blackbody with the corresponding effective
temperature or as a stellar atmosphere model with the appropriate
surface gravity and effective temperature. Here, we use blackbodies 
during the ionization cycles and to compute our MC
estimators. However, during the spectral synthesis stage of the 
simulation we use stellar atmosphere models. This produces more
realistic model spectra and allows us to test if recombination
emission from the wind base can fill in the Balmer jump, which is
always in absorption in these models. Our synthetic stellar atmosphere
spectra are calculated with
\textsc{Synspec}\footnote{http://nova.astro.umd.edu/Synspec43/synspec.html}
from either Kurucz \citep{kurucz1991} atmospheres (for $T_{eff} \leq
50,000$~K) or from \textsc{TLUSTY} models \citep{tlusty} (for $T_{eff} > 50,000$~K).

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/sed_figure.eps}
\caption{
The spectral energy distribution (SED) of the 
White Dwarf (dashed line) and blackbody accretion
disk used in the ionization calculation (black solid line).
The positions of the Hydrogen and Helium ionization edges 
are marked with vertical lines.
{\bf JM: do we want to show the stellar atmosphere spectrum here?
we know it looks a bit crap around the He II edge, so perhaps not...
 }
 }
\label{sed}
\end{figure}

@@

where $\rm{Q_H}$ is the number of hydrogen-ionizing photons emitted by
the illuminating source per second, $R$ is the distance to the source,
$\rm{n_H}$ is the local hydrogen number density and $c$ is the speed of
light. $J_\nu$ is the MC estimator for the mean intensity in the cell. 
The ionization parameter is a useful measure
of the ionization state of a plasma, as it evaluates the ratio 
of the integrated ionizing radiation field and the local Hydrogen density.

@@

The emissivity of these recombination 
features scales as $\rho^2$, meaning that they form almost entirely in the 
dense base of the wind, just above the accretion disk. Here, the
velocity field of the wind is still dominated by rotation, rather than
outflow, which accounts for the double-peaked shape of the lines. In
principle, lines formed in this region can still be single peaked,
since the existence of a poloidal velocity {\em gradient} changes the
local escape probabilities (MC96, MC97). However, as
demonstrated explicitly in section~\ref{modelb}, the line opacity in our
benchmark model is not high enough for this radiative transfer effect
to dominate the line shapes.

@@

In the UV, the model still produces all the observed lines, 
and deep P-Cygni profiles are produced in the normal resonance lines,
as discussed in section ??. However, the UV spectra also
display what is perhaps the biggest problem with this revised model,
namely the strength of resonance line emission 
at low and intermediate inclinations.
In order to generate strong optical wind signatures, we have adopted wind
parameters that lead to very high densities at the base of the wind
($n_e\sim10^{13}-10^{14}$ in the main emission region). This produces
the desired optical recombination emission, but also increases the
role of collisional excitation in the formation of the UV resonance
lines. This explains the pronounced increase in the emission component 
of the C\textsc{iv} 1550\AA\ resonance line, for example, relative to
what was seen in the benchmark model (compare figures~\ref{spec} and
\ref{uvoptb}). The strength of this component in the revised model 
is probably somewhat too high to be consistent with UV observations 
of high-state CVs (see e.g. Long et al. 1991, 1994; Noebauer et al. 2010).
\nocite{long1991,long1994, noebauer}

@@

% Although \ha\ is single-peaked in our model, the majority of the
% lines are double-peaked, particularly  
Figure~\ref{halpha} shows how the H$\alpha$
% He~{\sc ii}~$\lambda 4686$ 
line changes with the kinematics of the wind
for an inclination of $80^\circ$. 
% The width and strength of He~{\sc ii} 
% clearly depend on both $R_v$ and $\alpha$, but the overall line profile 
% shape is always double-peaked. This contrasts with the behaviour of 
The line becomes single-peaked in our densest models. 
The transition from double-peaked to single-peaked occurs when the 
line becomes optically thick. This reproduces the behaviour predicted by MC96, who
showed that the poloidal velocity gradient in the wind can modify the
shape of optically thick emission lines even in the near-disk regime,
where the poloidal velocity itself is much smaller than the rotational
velocity. This has the effect of suppressing the line wings and
can therefore lead to the production of a single-peaked line profile
shape.

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/mc.eps}
\caption{
\ha\ line profiles, normalised to 1, plotted in veliocity space 
for three models with varying kinematic 
properties, computed at an inclination of $80^\circ$.
The benchmark model and the improved optical
model described in section 6 are labeled as A and B respectively,
and a third model (X) which has an increased acceleration length of 
$R_V = 283.8~R_{WD}$, and $\alpha=4$ is also shown. 
The $x$-axis limits correspond to the Keplerian velocity at 
$4R_{WD}$, the inner edge of the wind.
We observe a transition from double to single-peaked lines
as the emission region becomes more optically thick.
% The parameter that has been altered from the benchmark model 
% is marked in each column and row. Models towards the bottom right
% are have more slowly accelerating winds and denser wind bases. 
% {\bf JM: Need to discuss the `single-peaked-ness' of line profiles
% in our telecon.}
}
\label{halpha}
\end{figure} %fullpage

@@

There are also interesting differences between the revised
model and the RW Tri data set. For example, the model exhibits
considerably stronger He~{\sc ii} features than the observations,
which suggests that the overall ionization state of the model is
somewhat too high. 
As discussed in section 5.3, H$\alpha$ is the only line
that is single-peaked in the model. This is
in contrast to what is seen in the observations,
although the relatively low resolution of the RW Tri
spectra makes this comparison difficult.

@@

We have investigated whether a disk wind model designed to reproduce
the UV spectra of high-state CVs would also have a significant effect
on the optical spectra of these systems. We find that this is indeed
the case. In particular, the model wind produces H and He
recombination lines, as well as a recombination continuum blueward of
the Balmer edge.

@@

\renewcommand{\labelitemi}{$\bullet$}
\begin{itemize}
	\item fill in the Balmer absorption edge in the spectrum of
          the accretion, thus accounting for the absence of a strong edge in
          observed CV spectra \citep{KLWB98};
	\item produce the commonly observed single-peaked optical line
          profiles (MC96)
\end{itemize}
\smallskip

Although we note that limb darkening effects on disk atmosphere opacities
would affect the Balmer jump. We can also make a number of basic,
testable predictions for wind-formed signatures:

>>>>

\begin{abstract}
Many high-state non-magnetic cataclysmic variables (CVs) exhibit
blue-shifted absorption or P-Cygni profiles in ultraviolet
(UV) resonance lines, such as C~{\sc iv}~1550\AA. These features
imply the existence of powerful accretion disk winds in
CVs.
% and considerable effort has gone into modelling UV line formation
% in such outflows. By contrast, much less attention has been paid to
% the potential impact of these disk winds on the {\em optical} spectra
% of CVs. 
Here, we use an improved version of our Monte Carlo ionization
and radiative transfer code to investigate whether disk wind models that
produce realistic UV line profiles are also likely to generate
observationally significant recombination line and continuum emission
in the {\em optical} waveband.   
In particular, we test whether outflows may 
be responsible for the single-peaked emission line profiles often seen
in high-state CVs and for the weakness of the Balmer absorption edge
(relative to simple models of optically thick accretion disks).
We find that a standard disk wind model that is successful in
reproducing the UV spectra of CVs also leaves a noticeable imprint on
the optical spectrum, particularly for systems viewed at high
inclination. The strongest optical wind-formed recombinations lines are \ha\ and
\heiiopt\, although the outflow also produces emission in the Ly-$\alpha$
and \heiiuv\ UV lines. However, all of the optical lines produced by
this default wind model are double-peaked, and the effect of the wind
on the Balmer jump is modest. 
We demonstrate that a higher-density outflow model 
displays reasonable verisimilitude with the optical spectrum of RW Trianguli,
producing all the observed lines. In addition, this model
produces recombination continuum sufficient to fill in the Balmer jump,
although it also predicts stronger-than-observed 
collisionally excited UV emission lines. 
Our revised model also significantly
alters the spectral shape of the continuum and displays
a strong inclination dependence in its emission properties. 
No single-peaked emission is seen, although we observe a 
narrowing of the double-peaked emission lines from the base of the wind. 
We suggest that a non-homogenous, `clumpy' wind may provide a possible
solution.
On the basis of these results, we suggest that winds can 
modify, and perhaps even dominate, the line and continuum emission
from accreting systems.
% We carry out a limited exploration of
% the disk wind parameter space, focusing on those parameters that affect the
% density near the base of the wind. This shows that higher-density
% outflows can produce both single-peaked emission lines and a
% recombination continuum sufficient to fill in the Balmer
% jump, although such models also tend to predict stronger-than-observed 
% collisionally excited UV emission lines. Our main conclusion is that
% disk wind emission is likely to have a significant impact on the
% optical spectra of high-state CVs. 
% {\bf JM: This abstract is a few lines longer than I would like- 
% I think something punchier is required.}
\end{abstract}

@@

It has been known for a long time that winds emanating from the
accretion disk are important in shaping the ultraviolet (UV) spectra
of high-state CVs \citep{heap1978, greensteinoke1982}. The most spectacular evidence for such
outflows are the P-Cygni-like profiles seen in UV resonance lines such as
\civfull\ (see e.g. Cordova \& Mason
1982\nocite{cordova1982}). Considerable effort has been spent over the
years on understanding and modelling these UV features (e.g. Drew \&
Verbunt 1985\nocite{drewverbunt1985}; Mauche \& Raymond
1987\nocite{maucheraymond1987}; Drew 1987; Shlosman \& Vitello 1993; [hereafter
SV93]\nocite{SV93}; Knigge, Woods \& Drew 1995 [hereafter
KWD95]\nocite{KWD95}; Knigge \& Drew 1997\nocite{kd1997}; 
Knigge et al. 1997\nocite{knigge1997}; Long \&
Knigge 2002 [hereafter LK02]\nocite{LK02}, Noebauer et al. 2010\nocite{noebauer};
Puebla et al. 2011\nocite{puebla2011}). The basic picture emerging from these efforts is
of a slowly accelerating, moderately collimated bipolar
outflow that carries away $\simeq 1\% - 10\%$ of the accreting
material. State-of-art-simulations of line formation in this type
of disk wind can produce UV line profiles that are remarkably similar
to observations.

@@

Murray \& Chiang (1996, 1997; hereafter referred to collectively as MC96)\nocite{MC96, MC97} 
have shown that the presence of disk winds may
offer a natural explanation for the single-peaked optical emission lines in
high-state CVs, since they can strongly affect the radiative transfer
of line photons. Strong support for a significant wind contribution to the
optical emission lines comes from observations of eclipsing
systems. There, the single-peaked lines are often only weakly
eclipsed, and a significant fraction of the line flux remains visible
even near mid-eclipse (REF). This points to line formation in a spatially
extended region, such as a disk wind (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence for a wind contribution to the optical lines comes
from isolated observations of P-Cygni-like line profiles even in optical
lines, such as \ha\ and He \textsc{i} $\lambda5876$ \citep{patterson1996, RN98, kafka2004}.

@@

\begin{figure}	%fullpage
\centering
\includegraphics[width=0.8\textwidth]{figures/fig1.eps}
\caption{Optical spectra of three nova-like variables: 
RW Sex (top; Beuermann et al. 1992),
IX Vel (top middle; Beuermann \& Thomas 1990) 
and RW Tri in and out of eclipse (bottom two panels; Groot et al. 2004).
The data have been digitized from the respective publications and a few of the more
prominent emission and absorption lines are marked. 
These systems have inclinations of $30^\circ$, $60^\circ$ and $70^\circ$ 
respectively. 
The trend of increasing Balmer line emission with inclination can be seen.
In RW Tri strong single-peaked emission in the Balmer lines is seen even
in eclipse, indicating that the lines may be formed in a spatially
extensive disk wind, and there is even a suggestion 
of a (potentially wind-formed) recombination continuum in the eclipsed
spectrum. We have attempted to show each spectrum over a similar dynamic range.
}
\label{novalikes}
\end{figure}

@@

\py is a Monte Carlo  ionization and radiative transfer code which
uses the Sobolev approximation to treat line transfer. It has already
been described extensively in LK02, SDL05 and H13, so here we provide
only a brief summary of its operation, focusing particularly on new
aspects of our implementation of macro-atoms into the code.

@@

\py operates in two distinct stages. First, the ionization state,
level populations and temperature structure are calculated. This is
done iteratively, by 
propagating several populations of Monte Carlo energy quanta (`photons')
through a model wind. The geometric and kinemetic properties of the
outflow are specified on a pre-defined spatial grid. In each of these
iterations (``ionization cycles''), the code records estimators that 
characterize the radiation field in each grid cell. At the end 
of each ionization cycles, a new electron temperature is calculated
that more closely balances heating and cooling in the 
plasma. The radiative estimators and updated electron
temperature are then used to revise the ionization state of the wind,
and a new ionization cycle is started. The process is repeated until
heating and cooling are balanced throughout the wind.

@@

Lucy (2002, 2003\nocite{lucy2002, lucy2003}; hereafter L02, L03) 
has shown that it is possible to calculate the emissivity of a gas in
statistical equilibrium accurately by quantising matter into
`macro-atoms', and radiant and kinetic energy into indivisible energy
packets (r- and k- packets, respectively). His macro-atom scheme
allows for all possible transition paths from a given level and
provides a full non-local thermodynamic equilibrium (NLTE) solution
for the level populations based on Monte Carlo estimators. The macro-atom
technique has already been used to model Wolf-Rayet star
winds \citep{sim2004}, AGN disk winds \citep{simlong2008, tatum2012},
supernovae \citep{kasen2006, kerzendorfsim} and YSOs (SDL05). A full 
description of the can be found in L02 and L03.

Briefly, macro-atom NLTE level populations and ionization fractions
are calculated by solving the statistical equilbrium equations between
each pair of levels. For example, the bound-bound de-excitation rate,
${\cal R}_{ul}$, in an ion is given by  
\begin{equation}
{\cal R}_{ul} = \beta_{lu} A_{ul} n_u + \beta_{lu} B_{ul} n_u J_{lu} +
C_{ul} n_u n_e,\label{eq:nlte_rul},
\end{equation}
where $u$ and $l$ denote the upper and lower levels, $C$ represents the
collisional rate coefficients and $A$ and $B$ represent the Einstein
coefficients. The quantity $\beta_{lu}$ is the probability that a
given line photon will escape the Sobolev region. Finally, $J$ is the
Monte Carlo estimator for the mean intensity in the line, as recorded during
the photon propagation. The corresponding excitation rate is then 
\begin{equation}
{\cal R}_{lu} = \beta_{lu} B_{lu} n_l J_{lu} + C_{lu} n_l n_e,
\end{equation}
where we implicitly include a correction for stimulated emission in
our definition of the estimate for $J_{ul}$.

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/acc_law.eps}
\caption{
The adopted poloidal velocity law for various values of the
acceleration exponent, $\alpha$.
{\bf JM: KSL has suggested we ditch this}
} 
\label{acc_law}
\end{figure}

@@

\py has has some flexibility when treating the accretion 
disk as a source of photons. The disk is broken down into annuli 
such that each annulus contributes an equal amount to the bolometric
luminosity. We take the disk to be geometrically thin, but optically
thick and thus adopt the temperature profile of a standard
\cite{shakurasunyaev1973} $\alpha$-disk. An annulus can then
be treated either as a blackbody with the corresponding effective
temperature or as a stellar atmosphere model with the appropriate
surface gravity and effective temperature. Here, we use blackbodies 
during the ionization cycles and to compute our Monte Carlo
estimators. However, during the spectral synthesis stage of the 
simulation we use stellar atmosphere models. This produces more
realistic model spectra and allows us to test if recombination
emission from the wind base can fill in the Balmer jump, which is
always in absorption in these models. Our synthetic stellar atmosphere
spectra are calculated with
\textsc{Synspec}\footnote{http://nova.astro.umd.edu/Synspec43/synspec.html}
from either Kurucz \citep{kurucz1991} atmospheres (for $T_{eff} \leq
50,000$~K) or from \textsc{TLUSTY} models \citep{tlusty} (for $T_{eff} > 50,000$~K).

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/sed_figure.eps}
\caption{
The spectral energy distribution (SED) of the 
White Dwarf (dashed line) and blackbody accretion
disk used in the ionization calculation (black solid line).
The positions of the Hydrogen and Helium ionization edges 
are marked with vertical lines.
{\bf JM: KSL has suggested we ditch this}
 }
\label{sed}
\end{figure}

@@

where $\rm{Q_H}$ is the number of hydrogen-ionizing photons emitted by
the illuminating source per second, $R$ is the distance to the source,
$\rm{n_H}$ is the local hydrogen number density and $c$ is the speed of
light. $J_\nu$ is the Monte Carlo estimator for the mean intensity in the cell. 
The ionization parameter is a useful measure
of the ionization state of a plasma, as it evaluates the ratio 
of the integrated ionizing radiation field and the local Hydrogen density.

@@

Our models do not exhibit P-Cygni profiles in the optical lines.
This is perhaps not surprising. LK02 and SV93 originally designed such models
to reproduce the UV line profiles. Thus, most of the wind
has an ionization parameter of $\log U \sim 2$ (see figure 6).
This means Hydrogen and Helium are fully ionized throughout 
much of the wind and is successful in producing recombination features.
However, it leads to absorbing columns in \ha\ of $N_H\sim ??$ - far too
low to produce noticeable blue shifted absorption. 
While we suspect that the winds that exhibit such profiles must 
lie in a different ionization regime, requiring 
significantly different disk SEDs or mass-loss rates to a 
`standard' high-state CV. However, we note it is difficult for this
idea to exist within a line-driven wind framework.

The emissivity of these recombination 
features scales as $\rho^2$, meaning that they form almost entirely in the 
dense base of the wind, just above the accretion disk. Here, the
velocity field of the wind is still dominated by rotation, rather than
outflow, which accounts for the double-peaked shape of the lines. In
principle, lines formed in this region can still be single peaked,
since the existence of a poloidal velocity {\em gradient} changes the
local escape probabilities (MC96). However, as
discussed further in section~\ref{modelb}, the 
radial velocity shear in our
models is not high enough for this radiative transfer effect
to dominate the line shapes.

@@

In the UV, the model still produces all the observed lines, 
and deep P-Cygni profiles are produced in the normal resonance lines,
as discussed in section 4.2. However, the UV spectra also
display what is perhaps the biggest problem with this revised model,
namely the strength of resonance line emission 
at low and intermediate inclinations.
In order to generate strong optical wind signatures, we have adopted wind
parameters that lead to very high densities at the base of the wind
($n_e\sim10^{13}-10^{14}$ in the main emission region). This produces
the desired optical recombination emission, but also increases the
role of collisional excitation in the formation of the UV resonance
lines. This explains the pronounced increase in the emission component 
of the C\textsc{iv} 1550\AA\ resonance line, for example, relative to
what was seen in the benchmark model (compare figures~\ref{spec} and
\ref{uvoptb}). The strength of this component in the revised model 
is probably somewhat too high to be consistent with UV observations 
of high-state CVs (see e.g. Long et al. 1991, 1994; Noebauer et al. 2010).
\nocite{long1991,long1994, noebauer}

@@

Figure~\ref{halpha} shows how the H$\alpha$
line changes with the kinematics of the wind
for an inclination of $80^\circ$. 
The line narrows as the models move towards denser,
more slowly accelerating winds. 
This is due
the base of the wind becoming optically thick such 
that line emission cannot escape. This means
what is effectively a photosphere for the line moves 
to a greater vertical distance from the disk plane,
where rotational velocities drop off as $1/r$. 
The narrowing of the line is {\em not} due to the radial velocity shear
mechanism proposed by MC96.
The main criterion for this mechanism is

\begin{equation}
\frac{dv_p}{dr} > \frac{dv_\phi}{ds}\,
\end{equation}

where $v_p$ is the poloidal velocity and $v_\phi$
is the Keplerian velocity.
This criterion is simply not satisfied in the emission region of our models.

There are two possible ways in which the model
could be altered to better reproduce single-peaked
lines as a result of radial velocity shear. First,
one could adopt a more equatorial wind geometry. 
This would increase the velocity gradient in the radial 
direction. However, CV winds are known to be moderately collimated.
This is demonstrated by the fact that blue-shifted resonance line 
absorption is seen in low inclination systems such as RW Sex \citep{greensteinoke1982},
but not in eclipsing systems \citep{cordovamason1985}.
Furthermore, the main reason for adopting this type of 
wind geometry is that it already successfully models the 
P-Cygni resonance lines seen in the UV (LK02, Noebauer et al. 2010),
We note that equatorial winds are expected in AGN and X-ray binaries 
\citep{ponti2012}, and that these winds may be more likely
to exhibit the required kinematics.
Second, one could adopt velocity laws with higher 
velocity gradients in the line emission region. 
However, this poses problems for a smooth biconical wind. 
We have already demonstrated that high densities are required 
to produce significant emission. In a smooth model, which
satisfies the mass continuity condition (see section 3.1), 
it is therefore difficult to reconcile 
these high densities with the high velocity gradients required.

We suggest that clumping
in CV winds provides a possible solution.
Non-homogeneity could allow higher density pockets 
of emission to exist with radial velocity shear still being high.
Line-driven winds are expected to be unstable (REFS),
and a number of implementations have been used in the stellar wind
community (REFS). We therefore hope that further investigation of
clumped winds could help solve the current `mystery' of single-peaked 
line emission in high inclination systems.

% Although \ha\ is single-peaked in our model, the majority of the
% lines are double-peaked, particularly  
% Figure~\ref{halpha} shows how the H$\alpha$
% % He~{\sc ii}~$\lambda 4686$ 
% line changes with the kinematics of the wind
% for an inclination of $80^\circ$. 
% % The width and strength of He~{\sc ii} 
% % clearly depend on both $R_v$ and $\alpha$, but the overall line profile 
% % shape is always double-peaked. This contrasts with the behaviour of 
% The line becomes single-peaked in our densest models. 
% The transition from double-peaked to single-peaked occurs when the 
% line becomes optically thick. This reproduces the behaviour predicted by fsingle-pe96, who
% showed that the poloidal velocity gradient in the wind can modify the
% shape of optically thick emission lines even in the near-disk regime,
% where the poloidal velocity itself is much smaller than the rotational
% velocity. This has the effect of suppressing the line wings and
% can therefore lead to the production of a single-peaked line profile
% shape.

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/mc.eps}
\caption{
\ha\ line profiles, normalised to 1, plotted in veliocity space 
for three models with varying kinematic 
properties, computed at an inclination of $80^\circ$.
The benchmark model and the improved optical
model described in section 6 are labeled as A and B respectively,
and a third model (X) which has an increased acceleration length of 
$R_V = 283.8~R_{WD}$, and $\alpha=4$ is also shown. 
The $x$-axis limits correspond to the Keplerian velocity at 
$4R_{WD}$, the inner edge of the wind.
We observe a narrowing of the lines
but this is not due to the radial velocity shear mechanism (see section 5.3
for a discussion).
% The parameter that has been altered from the benchmark model 
% is marked in each column and row. Models towards the bottom right
% are have more slowly accelerating winds and denser wind bases. 
% {\bf JM: Need to discuss the `single-peaked-ness' of line profiles
% in our telecon.}
}
\label{halpha}
\end{figure} %fullpage

@@

There are also interesting differences between the revised
model and the RW Tri data set. For example, the model exhibits
considerably stronger He~{\sc ii} features than the observations,
which suggests that the overall ionization state of the model is
somewhat too high. 
As discussed in section 5.3, the optical lines are
narrow. This is in contrast to what is generally seen in observations
of NLs, although the relatively low resolution of the RW Tri
spectra makes a specific comparison difficult.

@@

We have investigated whether a disk wind model designed to reproduce
the UV spectra of high-state CVs would also have a significant effect
on the optical spectra of these systems. We find that this is indeed
the case. In particular, the model wind produces H and He
recombination lines, as well as a recombination continuum blueward of
the Balmer edge. We do not produce P-Cygni profiles
in the optical H and He lines, 
which are seen in a small fraction of CV optical spectra.
Possible reasons for this are briefly discussed in section 
4.2.

@@

\renewcommand{\labelitemi}{$\bullet$}
\begin{itemize}
	\item fill in the Balmer absorption edge in the spectrum of
          the accretion, thus accounting for the absence of a strong edge in
          observed CV spectra \citep{KLWB98};
	\item produce the optical lines observed in CVs.
\end{itemize}
\smallskip

Although we note that limb darkening effects on disk atmosphere opacities
would affect the Balmer jump (see section 5.2). In section 5.3, we demonstrate that
although single-peaked emission can be formed in our densest models, 
it is not due to the radial velocity shear mechanism proposed by MC96.
The emission in our models which produce single-peaked lines comes from
approximately $10^{11}$~cm above the disk plane, which is not supported
by eclipse-mapping observations (REFS). We suggest
that non-homogenous, 
`clumpy' line-driven winds may offer a potential solution.
